Updated Criteria for the Diagnostic Procedure for Parkinson’s Disease Dementia on Level I and their Validity in Deep Brain Stimulation Cohort

### 2025-05-14 _manuscript/manuscript.qmd
CHANGEEESSS!!!

Parkinson’s disease (PD) is a neurodegenerative disorder. It is primarily defined by a gradual onset of motor impairment, including rigidity, bradykinesia, postural instability and resting tremor. Additionally, patients frequently suffer from a range of non-motor impairments [@postuma2015], particularly cognitive decline. This state might result in Parkinson’s disease dementia (PDD) in a subset of patients [@meireles2012].

@@

Ethical approval for the study protocol was obtained from the Ethics Committee of the General University Hospital in Prague. Informed consent was secured from all patients prior to their neuropsychological assessments, in adherence to ethical research guidelines.

@@

Cognitive performance was evaluated at both Level I (abbreviated assessment) and Level II (comprehensive assessment) according to the standard MDS neuropsychology battery for Parkinson's Disease Mild Cognitive Impairment (PD-MCI) [@litvan2012, @bezdicek2017]. Level I was assessed using the Mini-Mental State Examination (MMSE) [@stepankova2014; @folstein1975] and the Montreal Cognitive Assessment (MoCA) [@kopecek2016; @nasreddine2005]. The neuropsychological assessment at Level II covered five cognitive domains, each evaluated through specific tests as follows: attention and working memory assessed using Trail Making Test Part A (TMT-A) [@bezdicek2012; @reitan2004], and WAIS Digit Span Backward (WAIS DSB) [@wechsler1997]; executive function evaluated via Categorical Verbal Fluency (CF) [@nikolai2015], and subtest from the Prague Stroop Test – Colors (PST-C) [@bezdicek2021]; language measured with the WAIS Similarities subtest [@wechsler1997], and the Boston Naming Test (BNT-60) [@zemanova2016; @kaplan1983]; memory examined using the Rey Auditory Verbal Learning Test (RAVLT) [@frydrychova2018; @bezdicek2013; @rey1964] delayed recall, and the Brief Visual Memory Test–Revised (BVMTR) [@havlik2020; @benedict1997] delayed recall, or WAIS Family Pictures subtest [@wechsler1997] delayed recall; visuospatial function assessed through the Judgment of Line Orientation Test (JoLO) [@benton1983], and Clock Drawing Test (CLOX) [@royall1998].

To assess functional impairment, the Functional Activities Questionnaire (FAQ) [@bezdicek2016; @pfeffer1982] was administered. Additionally, neuropsychiatric status was evaluated using the Beck Depression Inventory-II (BDI-II) [@ciharova2020; @beck1996] and State-Trait Anxiety Inventory (STAI) [@spielberg1983; @mullner1980]. Psychotic symptoms were assessed through structured psychiatric interviews conducted by a trained psychiatrist.

@@

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD algorithm listed in @tbl-crits (see also @tbl-algos) resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (algorithms) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an algorithm (column)[^6]. PDD rate estimates were computed as $\frac{N_{PDD}}{N_{total}}$ separately for each algortihm. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each algorithm for probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

[^6]: **JM: This could and should be shared most likely, as long, as we anonymize properly. Let's ask Oto Mestek and the NPO team how and if is it possible.**

@@

Data wrangling and visualizations were done in the *tidyverse* package [@wickham2019] and tables were formatted in the *gt* package [@iannone2024]. All analyses were conducted within the R (version `r with(version, paste(major, minor, sep="."))`) software environment for statistical computing [@rsoft]. The software code supporting this article is available at <https://github.com/josefmana/DemCr1t.git>.[^7]

[^7]: **JM: Do not forget to make it public before submitted!**

@@

Results of the analyses of prediction Accuracy, Cohen's $\kappa$, Sensitivity and Specificity are presented in @fig-acc, @fig-kappa, @fig-sens and @fig-spec respectively. **Numerical results are available at ...** [^8]. Generally, algorithms that employed the same operationalization of IADL deficit showed substantial pairwise concordance, however, algorithms that operationalized IADL deficit differently did not. Whereas among algorithms with identical IADL deficit operationalization, the agreement judged by Cohen's $\kappa$ was moderately high (operationalization by FAQ total score: $\kappa$ = `r kappa_sums["faq_tot"]`; operationalization by FAQ item 9: $\kappa$ = `r kappa_sums["faq_9"]`), among algorithms that differ in IADL deficit operationalization it was low ($\kappa$ = `r kappa_sums["cross"]`).

[^8]: **JM: Should provide an excel sheet or html table the readers could go through by themselves.**

@@

This study examined the diagnostic variability and validity of Level I criteria for PDD, **specifically in candidates for DBS treatment**[^9]. By systematically comparing 68 diagnostic algorithms based on MMSE, MoCA, sMoCA, and Level II criteria, our findings highlight the significance of algorithm design.

[^9]: **JM: All-right, I do not think it needs to be here either.**

@@

Our results showed a wide range in estimated PDD rate across algorithms, ranging from 2.00% to 16.75%. Estimates reached lower rates when using solely FAQ item 9 (as an approximation of the pill questionnaire suggested by @dubois2007) in comparison with the full FAQ scale. This discrepancy highlights the diagnostic importance of IADL assessment. Our overall PDD rates were consistently lower compared to previous studies regarding PDD among PD patients, **demonstrating wide variability based on various criteria used**[^10]. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigation found even higher rates, reaching up to 30% [@aarsland2005][^11]. A recent complex meta-analysis synthesizing global data placed the expected PDD rate in PD at 26.30% [@sousa2022]. **The discrepancy is likely attributable to our DBS-specific cohort, which differs in important characteristics, such as the younger age of the patients [@rana2012]. / Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in sample characteristics, diagnostic criteria, and methodology.** [^12] [^13]

[^10]: **JM: What is demonstrating this variability? Our findings or previous ones?**

[^11]: **JM: There was aarsland2011 originally but I did not find such a study. Is this reference correct?**

[^12]: **Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events**

[^13]: \*\*JM: I prefer the second one with added sentence like: "Specifically, our cohort included candidates for DBS who are generally younger than the general PD population and age was repeatedly shown to be a strong predictor of PDD (some citations)." We should also incorporate here a statement about missing evidence of age being predictor in our sample.\*

@@

Our pairwise comparisons of diagnostic algorithms showed that agreement was notably stronger among those using the same IADL operationalization **(**$\kappa$ range from 0.75, SD = 0.13 to 0.86, SD = 0.09)[^14] compared to those using different IADL definitions **(**$\kappa$ = 0.34, SD = 0.08)[^15]. **Moreover, the agreement was slightly higher between algorithms that defined IADL deficit by FAQ item 9 compared to algorithms that defined it using the full FAQ scale. One possible explanation of this difference follows from the observation that algorithms using the full scale definition yielded higher PDD rate estimates. Because there was higher probability of being diagnosed with IADL deficit based on the full FAQ scale, there was also a bigger room for disagreement in the Cognitive Impairment status when different indexes weer used (e.g., by defining executive deficit via clock drawing vs. lexical fluency).**

[^14]: **JM: There is no need to repeat these numbers in this section.**

[^15]: **JM: Same as above.**

**Overall, algorithms using the same definition of IADL deficit showed degree of concordance equivalent to moderate (when using FAQ total score to define IADL deficit) or strong (when using FAQ item 9 to define IADL deficit) level of agreement between raters in an inter-rater reliability analysis [@mchugh2012]. On the other hand, the concordance between algorithms using different IADL deficit definitions was equivalent to minimal agreement.** This demonstrates that even slight methodological differences can yield divergent diagnostic outcomes. Such findings are critical for clinicians relying on Level I criteria for eligibility decisions, as the choice of algorithm could lead to contradictory classifications of PDD.

@@

When using Level II diagnosis as the reference, MoCA-based Level I algorithms, particularly those using Clock Drawing to assess executive function, demonstrated the highest predictive accuracy. This supports recent proposals to modernize PDD diagnostic frameworks [@kulisevsky2024], favoring MoCA-derived components and broader functional assessment tools. In contrast, MMSE-based algorithms consistently underperformed, suggesting limited sensitivity in capturing the nuanced cognitive deficits characteristic of PDD.

The advantage of MoCA-based algorithm was evident when using the FAQ total score to define functional impairment in comparison with using solely FAQ item 9. In these cases, the best-performing algorithms achieved accuracy rates of 94–99%, together with high sensitivity and specificity. **JM: Some sMoCA praise should be added here ...**

## Clinical Implications^[**JM: Is this section needed? It does not seem to say much to me. I would't be afraid to drop it.**]

Our findings have immediate implications for DBS candidacy assessment. Given that a diagnosis of PDD typically excludes surgical intervention, inconsistent application of diagnostic criteria can lead to unbalanced access to the treatment.

Moreover, the relative success of multidomain assessments, such as Level II protocols, highlights the importance of psychometric precision. This approach may serve as a confirmatory tool in cases where Level I assessments provide ambiguous results.

@@

This study’s generalizability is limited by its retrospective design and the homogeneity of the patient cohort, which may not reflect broader PD populations with varying cognitive profiles. **Specifically, younger age of the sample and possible lack of patients with high risk phenotypes for PDD compared to general population of people with PD might have been responsible for comparatively lower PDD rate estimates in our study. Nonetheless, the analysis of the concordance between different algorithms might not be affected by the sample to the same degree allowing for a broader generalization. Future studies could replicate our results on different cohorts of PD patients. To make this process easier, the code used to generate our results is publicly available and easily applicable to other data sets with a similar data structure.**

**JM: The fact that no biomarkers were used and the IADL was at best estimated by FAQ which is somewhat archaic (comp. to the e-IADL thing which could be also cited and is more in line with Kulisevsky) should be mentioned. Also, this is a great space for an advert for NOT/DOT because the self-report of IADLs can be viewed as a limitation worth of addressing, especially since our results imply possible central role for IADL deficit definition in further work on PDD definition to achieve good reliability.**

Future research should focus on the validation of updated Level I criteria using longitudinal data and neurobiological markers. Furthermore, the development of adaptive diagnostic algorithms that can combine performance across cognitive and functional domains may enhance diagnostic sensitivity while reducing the risk of **over-classification of PDD**^[**JM: What's an over-classification of PDD?**].

@@

This study systematically investigated the application of multiple Level I diagnostic criteria for PDD within a cohort of patients considered for DBS treatment. Our results show variability in PDD rate estimates, strongly influenced by the choice of cognitive screening instruments and the operationalization of functional impairment. The divergence observed across algorithms demonstrates the sensitivity of diagnostic outcomes to seemingly negligible methodological choices.^[**JM: This sounds as a great opening to the discussion, rather than conclusions :-)**]

The **proposed changes**^[**JM: They did not propose anything though and from what I gather, the authors stress this a lot - that their article in not a proposal of new criteria, only a position paper.**] of the current diagnostic criteria [@kulisevsky2024] offer heightened sensitivity by proposing the use of MoCA-based components and broader IADL assessments. Conservative criteria, such as reliance on pill questionnaire (i.e. FAQ item 9 equivalence), may fail to detect meaningful functional decline and thus under-identify true cases of PDD. **On the other hand, such criteria show slightly higher agreement between their different forms.**

Conclusively, this study contributes to the improvement of PDD diagnostics by offering **evidence for the clarification of Level I criteria**^[**JM: What is meant by clarification of Level I criteria? Also, how does our study offer evidence for such thing?**] and **emphasizing the value of psychometric precision in clinical neuropsychology**^[**JM: This is independent of our study.**]. Future research should validate the algorithms longitudinally and explore possible integration of biomarkers to further improve diagnostic reliability.

@@

Estimates relating to RO1.1 and RO1.3 cannot be safely generalized beyond a population of PD patients that are candidates for DBS due to the systematic differences between DBS candidates pool and general PD population (such as the lower age of DBS candidates compared to the general PD population). On the other hand, the estimates relating to RO1.4 (and to a lesser degree to RO1.2[^16]) may not be substantially influenced by the sample at hand as the primary source of their variance might come from variability in measures employed (e.g., MMSE vs MoCA to assess global cognitive performance) rather than variability in patients' performance. Assuming that there is no substantial Differential Item Functioning for DBS candidates compared to a broader population of patients with PD, the estimates relating to RO1.4 can be cautiously generalized beyond the current sample.

[^16]: Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events

>>>>

Parkinson’s disease (PD) is a neurodegenerative disorder typically characterized by a progressive onset of motor symptoms, including rigidity, bradykinesia, postural instability and resting tremor. Moreover, patients suffer from a range of non-motor impairments [@postuma2015], particularly cognitive decline. This factor might result in Parkinson’s disease dementia (PDD) in a subset of patients [@meireles2012].

@@

@@

Cognitive performance was evaluated at both Level I and Level II according to the standard MDS battery for Parkinson's Disease Mild Cognitive Impairment (PD-MCI) [@litvan2012, @bezdicek2017]. Cognitive performance at Level I was assessed by the Mini-Mental State Examination (MMSE) [@stepankova2014; @folstein1975] and the Montreal Cognitive Assessment (MoCA) [@kopecek2016; @nasreddine2005]. The comprehensive neuropsychological assessment at Level II evaluated five cognitive domains through specific test: attention and working memory assessed by Trail Making Test Part A (TMT-A) [@bezdicek2012; @reitan2004], and WAIS Digit Span Backward (WAIS DSB) [@wechsler1997], executive functions by Categorical Verbal Fluency (CF) [@nikolai2015], and subtest from the Prague Stroop Test – Colors (PST-C) [@bezdicek2021], language by the WAIS Similarities subtest [@wechsler1997], and the Boston Naming Test (BNT-60) [@zemanova2016; @kaplan1983], memory by the Rey Auditory Verbal Learning Test (RAVLT) [@frydrychova2018; @bezdicek2013; @rey1964] delayed recall, and the Brief Visual Memory Test–Revised (BVMTR) [@havlik2020; @benedict1997] delayed recall, or WAIS Family Pictures subtest [@wechsler1997] delayed recall, visuospatial function assessed by the Judgment of Line Orientation Test (JoLO) [@benton1983], and Clock Drawing Test (CLOX) [@royall1998].

The Functional Activities Questionnaire (FAQ) [@bezdicek2016; @pfeffer1982] was administered to assess functional impairment. The Beck Depression Inventory-II (BDI-II) [@ciharova2020; @beck1996] and State-Trait Anxiety Inventory (STAI) [@spielberg1983; @mullner1980] were used to assess neuropsychiatric status. Psychotic symptoms were assessed through structured psychiatric interviews conducted by a trained psychiatrist.

@@

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD algorithm listed in @tbl-crits (see also @tbl-algos) resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (algorithms) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an algorithm (column)[^1]. PDD rate estimates were computed as $\frac{N_{PDD}}{N_{total}}$ separately for each algortihm. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each algorithm for probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

[^1]: **JM: This could and should be shared most likely, as long, as we anonymize properly. Let's ask Oto Mestek and the NPO team how and if is it possible.**

@@

Data wrangling and visualizations were done in the *tidyverse* package [@wickham2019] and tables were formatted in the *gt* package [@iannone2024]. All analyses were conducted within the R (version `r with(version, paste(major, minor, sep="."))`) software environment for statistical computing [@rsoft]. The software code supporting this article is available at <https://github.com/josefmana/DemCr1t.git>.[^2]

[^2]: **JM: Do not forget to make it public before submitted!**

@@

Results of the analyses of prediction Accuracy, Cohen's $\kappa$, Sensitivity and Specificity are presented in @fig-acc, @fig-kappa, @fig-sens and @fig-spec respectively. **Numerical results are available at ...** [^3]. Generally, algorithms that employed the same operationalization of IADL deficit showed substantial pairwise concordance, however, algorithms that operationalized IADL deficit differently did not. Whereas among algorithms with identical IADL deficit operationalization, the agreement judged by Cohen's $\kappa$ was moderately high (operationalization by FAQ total score: $\kappa$ = `r kappa_sums["faq_tot"]`; operationalization by FAQ item 9: $\kappa$ = `r kappa_sums["faq_9"]`), among algorithms that differ in IADL deficit operationalization it was low ($\kappa$ = `r kappa_sums["cross"]`).

[^3]: **JM: Should provide an excel sheet or html table the readers could go through by themselves.**

@@

This study systematically investigated the application of multiple Level I diagnostic criteria for PDD. Our results show variability in PDD rate estimates, strongly influenced by the choice of cognitive screening instruments (MMSE, MoCA and sMoCA) and the operationalization of functional impairment. The divergence observed across algorithms demonstrates the sensitivity of diagnostic outcomes to seemingly negligible methodological choices.

@@

Our results showed a wide range in estimated PDD rate across algorithms, ranging from 2.00% to 16.75%. Estimates reached lower rates when using solely FAQ item 9 (as an approximation of the pill questionnaire suggested by @dubois2007) in comparison with the full FAQ scale. This discrepancy highlights the diagnostic importance of how IADLs are assessed.

Our overall PDD rates were consistently lower than previous studies regarding PDD among PD patients, demonstrating wide variability based on various criteria used. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigation found even higher rate, reaching up to 30% [@aarsland2005]. A recent complex meta-analysis synthesizing global data placed the expected PDD rate in PD at 26.30% [@sousa2022]. Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in diagnostic criteria, methodology and sample characteristics. Specifically, age was repeatedly shown to be a strong predictor of PDD across studies[@rana2012; @sousa2022]. In our sample, we did not observe age-related PDD rate within our cohort, likely due to low age variability.

@@

Pairwise comparisons of diagnostic algorithms showed that agreement was notably stronger among those using the same IADL operationalization compared to those using different IADL definitions. Moreover, the agreement was slightly higher between algorithms that defined IADL deficit by FAQ item 9 compared to algorithms that defined it using the full FAQ scale. One possible explanation of this difference follows from the observation that algorithms using the full-scale definition yielded higher PDD rate estimates. Because there was a higher probability of being diagnosed with IADL deficit based on the full FAQ scale, there was also a bigger room for disagreement in the Cognitive Impairment status when different indexes were used (e.g. by defining executive deficit via clock drawing vs. lexical fluency).

Overall, when the same IADL definitions were used across algorithms, we observed concordance levels varying from moderate (using FAQ total score) to strong (using FAQ item 9), consistent with inter-rater reliability analysis [@mchugh2012]. Contrarily, the concordance between algorithms using different IADL deficit definitions was equivalent to minimal agreement. This demonstrates that even slight methodological differences can yield divergent diagnostic outcomes. Such findings are critical for clinicians relying on Level I criteria for eligibility decisions, as the choice of algorithm could lead to contradictory classifications of PDD.

@@

Using Level II diagnosis as the reference, MoCA-based Level I algorithms, particularly those using Clock Drawing to assess executive function, demonstrated the highest predictive accuracy. This supports recent proposals to modernize PDD diagnostic frameworks [@kulisevsky2024], favouring MoCA-derived components and broader functional assessment tools. In contrast, MMSE-based algorithms consistently underperformed, suggesting limited sensitivity in capturing cognitive deficits typical in PDD.

The advantage of MoCA-based algorithm was evident in combination with using the FAQ total score, reaching accuracy rates of 94–99%, along with high sensitivity and specificity. Furthermore, in the algorithm using sMoCA, the raw accuracy was moderate, however, the balanced accuracy (i.e. combined sensitivity and specificity) was robust. These findings imply that simplified tools such as sMoCA may be a promising tool for heightening efficiency of the assessment while maintaining diagnostic accuracy.

@@

This study’s generalizability is limited by its retrospective design and the homogeneity of the patient cohort, which may not reflect broader PD populations with varying cognitive profiles. Specifically, the younger age of the sample and possible lack of high-risk phenotypes for PDD compared to the general population of people with PD might have been responsible for comparatively lower PDD rate estimates in our study.

Nevertheless, the analysis of the algorithm concordance might not be affected by the sample to the same degree allowing for a broader generalization. Future studies should aim to replicate our results on larger and different cohorts of PD patients. To make this process easier, the code used to generate our results is publicly available and easily applicable to similarly structured data.

Furthermore, another limitation is the use of the FAQ questionnaire for IADL assessment. FAQ is highly subjective and informant-reliant, which may result in biased data. We recommend the use of more objective, modernized tools for IADl assessment, such as performance-based methods[@schmitter-edgecombe2020] or questionnaire adaptation including questions regarding gadgets use and digital literacy [@postema2024]. Our results emphasise the importance of IADL deficit for the PDD diagnosis, therefore, we advocate for using more reliable tools with high ecological validity.

@@

Our study highlights the variability in PDD classification across Level I diagnostic algorithms, heavily influenced by IADL operationalization and the choice of cognitive screening tools. The findings support the call for a change of the current diagnostic criteria [@kulisevsky2024], favouring the use of MoCA-based components and comprehensive IADL assessments.

Conservative criteria, such as reliance on pill questionnaire (i.e. FAQ item 9 equivalence), may fail to detect functional decline and thus under-identify true cases of PDD. Importantly, concordance across algorithms rises significantly, reaching moderate to high values, when the same definition of IADL is used (either FAQ total or FAQ item 9). Moreover, when using MoCA-based algorithms instead of MMSE-based ones, we can observe better approximations to the Level II battery.

Future research should validate these strategies longitudinally and explore the possible integration of biomarkers, neuropsychiatric variables (e.g. anxiety profile) and modernised IADL methods to further improve diagnostic reliability, accuracy and consistency of PDD diagnostics.

@@

Estimates relating to RO1.1 and RO1.3 cannot be safely generalized beyond a population of PD patients that are candidates for DBS due to the systematic differences between DBS candidates pool and general PD population (such as the lower age of DBS candidates compared to the general PD population). On the other hand, the estimates relating to RO1.4 (and to a lesser degree to RO1.2[^4]) may not be substantially influenced by the sample at hand as the primary source of their variance might come from variability in measures employed (e.g., MMSE vs MoCA to assess global cognitive performance) rather than variability in patients' performance. Assuming that there is no substantial Differential Item Functioning for DBS candidates compared to a broader population of patients with PD, the estimates relating to RO1.4 can be cautiously generalized beyond the current sample.

[^4]: Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events
